---
title: "Visualization of metabolites using rWikiPathways and RCy3"
auther: "ken kamiya"
date: "2019/07/07"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Abstract
This is the support page for "メタボロームデータ解析および解釈に資する可視化手法" in CICSJ Bulletin July 2019.
This page describes the script in detail regarding the last chapter of pathway visualization.

#Data
Datasets and assets I used are here.
 - Pathway 
   AtMetExpress Overview - Wikipathways https://www.wikipathways.org/index.php/Pathway:WP3622

 - Metabolite data
  AraMetLeaves - attached to DiffCorr package
  The Arabidopsis metabolome of the aerial regions of individual WT plants and the mto1 and tt4 mutants were analyzed by GC-TOF/MS.
  mto1 shows strong methionine accumulation.
   - DiffCorr: https://cran.r-project.org/package=DiffCorr
   - Reference of metabolite data: M. Kusano et al, (2007) https://doi.org/10.1186/1752-0509-1-53


#Step 0: Intstall Cytoscape
If you haven't yet installed Cytoscape, install it before running this script.
https://cytoscape.org/download.html
The following script only works if you have launched Cytoscape.

#Step 1: Prepare R Packages and Metabolite Data
This pattern of installation had succeeded on my environment(Ubuntu 16.04, R 3.6.0).
It doesn't matter how you do it, so install "rWikiPathways", "RCy3", "MetabAnalystR" according to your own environment.

```{r prepare R packages}
install.packages("BiocManager")
BiocManager::install("rWikiPathways")
library(rWikiPathways)
BiocManager::install("RCy3")
library(RCy3)

#Prepare MetaboAnalystR
install.packages("pacman")
library(pacman)
pacman::p_load(Rserve, ellipse, scatterplot3d, Cairo, randomForest, caTools, e1071, som, impute, pcaMethods, RJSONIO, ROCR, globaltest, GlobalAncova, Rgraphviz, preprocessCore, genefilter, pheatmap, SSPA, sva, Rcpp, pROC, data.table, limma, car, fitdistrplus, lars, Hmisc, magrittr, methods, xtable, pls, caret, lattice, igraph, gplots, KEGGgraph, reshape, RColorBrewer, tibble, siggenes, plotly, xcms, CAMERA, fgsea, MSnbase, BiocParallel, metap, reshape2, scales)
install.packages("devtools")
library(devtools)
devtools::install_github("xia-lab/MetaboAnalystR", build = TRUE, build_opts = c("--no-resave-data", "--no-manual", "--no-build-vignettes"))
library(MetaboAnalystR)
```
```{r prepare metabolite data}
install.packages("DiffCorr")
library(DiffCorr)
data("AraMetLeaves") #data for visualization

Col0.index <- grep("Col0", colnames(AraMetLeaves), perl = TRUE)
mto1.index <- grep("mto1", colnames(AraMetLeaves), perl = TRUE)
Col0Mean <-  apply(AraMetLeaves[, Col0.index], 1, mean)
mto1Mean <-  apply(AraMetLeaves[, mto1.index], 1, mean)
```

#Step 2: Gather informations (espescially metabolite ID)

##Get graphIds from the pathway
Before getting pathways you need to know "graphId" in WikiPathways.
https://www.wikipathways.org/index.php/Help:WikiPathways_Webservice/API#GraphId
graphId is an individual ID of a metabolite assigned in the pathway. By using this, the pathway metabolite and metabolite data are linked.

```{r get graphids from pathway}

pathway.HMDBIDs <- rWikiPathways::getXrefList(pathway="WP3622", systemCode="Ch")
pathway.graphIds <-  NULL
for(i in 1:length(pathway.HMDBIDs)){
  pathways <- rWikiPathways::findPathwaysByXref(pathway.HMDBIDs[[i]], systemCode="Ch") 
  targeted.pathway <- pathways[purrr::map_lgl(pathways, ~ .$id == "WP3622")][[1]]
  pathway.graphIds[[i]] <- as.character(targeted.pathway$fields$graphId$values)
}
```

##Get IDs of metabolite data using MetaboAnalystR
Metabolite data names are written in various ways and are not stable.
So we convert names to IDs such as HMDB or KEGG using MetaboAnalystR.
The same can be done with MetaboAnalyst (https://metaboanalyst.ca) on the Web. Entering a metabolite name returns the match result recognized by MetaBoAnalyst, so you can absorb minor differences in metabolite names.
In addition, MetaboAnalyst can analyze while generating R code, so it is also possible to use the displayed R code in MetaBoAnalystR after performing ID conversion manually with MetaboAnalyst on the Web.

```{get ids of metabolite data using metaboanalystr}
mSet<-InitDataObjects("NA", "utils", FALSE)
cmpd.vec<- rownames(AraMetLeaves)
mSet<-Setup.MapData(mSet, cmpd.vec);
mSet<-CrossReferencing(mSet, "name", T, T, T, T, T);
mSet<-CreateMappingResultTable(mSet)
mSet<-PerformDetailMatch(mSet, "Inositol-1-phosphate");
mSet<-GetCandidateList(mSet);
mSet<-SetCandidate(mSet, "Inositol-1-phosphate", "Inositol phosphate");
mSet<-PerformDetailMatch(mSet, "D-Glucose-6-phosphate");
mSet<-GetCandidateList(mSet)
mSet<-SetCandidate(mSet, "D-Glucose-6-phosphate", "Beta-D-Glucose 6-phosphate")
```

##Get metabolites with graphId in the pathway

The metabolites having pathwayId in the metabolite data are extracted.

```{r get metabolites with graphId in the pathway}
mSet.map.table <- mSet$dataSet$map.table
mSet.graphIds <- NULL
for(i in 1:length(mSet.map.table[,"HMDB"])){
  pathways <- findPathwaysByXref(mSet.map.table[,"HMDB"][[i]], systemCode="Ch")
  targeted.pathway <- pathways[purrr::map_lgl(pathways, ~ .$id == "WP3622")]
  if(length(targeted.pathway) == 0){
    mSet.graphIds[[i]] <- "NA"
  }
  if(!length(targeted.pathway) == 0){
    mSet.graphIds[[i]] <- targeted.pathway[[1]]$fields$graphId$values
  }
}
```

##Calculate log2FC from Col0 and mto1
For visualization, log2FC is calculated from the mean of Col0 and the mean of mto1 mutants.

```{r calculate log2fc}
log2FC <-  log2(mto1Mean / Col0Mean)
min.log2FC <-  min(log2FC, na.rm = TRUE)
max.log2FC <-  max(log2FC, na.rm = TRUE)
abs.log2FC <-  max(abs(min.log2FC), max.log2FC)
data.values <-  c(-abs.log2FC, 0, abs.log2FC)
```

## Make final matrix
Create a matrix to pass to Cytoscape.

```{r make matrix}
mSet.map.table.graphId.log2FC <-  data.frame(mSet.map.table, mSet.graphIds, log2FC)
```

#Step 3: Connect to Cytoscape and Visualize

Connect to Cytoscape using RCy3 for visualization.
**Start Cytoscape before running this chapter**

```{r connect to cytoscape and visualize}
#Install Wikipathways addin to Cytoscape
installApp('WikiPathways')  #only available in Cytoscape 3.7.0 and above

#Connect to Cytoscape
cytoscapePing()

#Pass the pathway to Cytoscape
RCy3::commandsRun('wikipathways import-as-pathway id=WP3622') 
toggleGraphicsDetails()

#load metabolite data and visualize 
loadTableData(mSet.map.table.graphId.log2FC, data.key.column = "mSet.graphIds", table.key.column = "GraphID")
node.colors <- c(rev(brewer.pal(length(data.values), "RdBu")))
setNodeColorMapping("log2FC", data.values, node.colors, default.color = "#FFFFFF", style.name = "WikiPathways")

#emphasize methionine
setNodeBorderColorBypass(node.names = "Methionine", new.colors = "#FF0000")
setNodeBorderWidthBypass(node.names = "Methionine", new.sizes = 10)
```

